---
title: "Exploratory Analysis of DirectSampling"
author: "Nicolàs Merino"
date: "2024-10-16"
output:
  html_document:
    theme: flatly
    toc: yes
    toc_depth: 5
    number_sections: true
    toc_float: 
      collapsed: false
    bookdown: yes
    highlight: tango
---

```{r setup, include=FALSE}
library(here)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here())
```

# Libraries
```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(corrplot)
library(GGally)
library(readr)
library(here)
```
# Data
```{r, warning=FALSE, message=FALSE}
# Load the data
df <- read_csv("drago_result_with_herbivory_new.csv")
```
```{r, warning=FALSE}
# Remove simulations with NaNs
df <- subset(df, complete.cases(df))
str(df)
```

# Descriptive Statistics
## For parameters
```{r}
param_stats <- df %>%
  select(sigma, epsilon, alfa, sigma_comp, assymetry) %>%
  summary()
print(param_stats)
```

## For metrics
```{r}
metric_stats <- df %>%
  select(avg_shannon, avg_bbp, alive_predators, mean_tl, mean_n_of_species) %>%
  summary()
print(metric_stats)
```

# Distribution Analysis
## Histograms for parameters
```{r}
param_names <- c("sigma", "epsilon", "alfa", "sigma_comp", "assymetry")
for (param in param_names) {
  p <- ggplot(df, aes(x = .data[[param]])) +
    geom_histogram(
      binwidth = (max(df[[param]], na.rm = TRUE) - min(df[[param]], na.rm = TRUE)) / 100,
      fill = "blue", color = "black", alpha = 0.7, boundary = 0
    ) +
    theme_minimal() +
    labs(title = paste("Histogram of", param), x = param, y = "Frequency")
  print(p)
}
```

## Histograms for metrics
```{r}
metric_names <- c("avg_shannon", "avg_bbp", "alive_predators", "mean_tl", "mean_n_of_species")
for (metric in metric_names) {
  p <- ggplot(df, aes(x = .data[[metric]])) +
    geom_histogram(
      binwidth = (max(df[[metric]], na.rm = TRUE) - min(df[[metric]], na.rm = TRUE)) / 30,
      fill = "green", color = "black", alpha = 0.7, boundary = 0
    ) +
    theme_minimal() +
    labs(title = paste("Histogram of", metric), x = metric, y = "Frequency")
  print(p)
}
```

# Correlation Matrix
```{r}
# Select numeric variables
numeric_vars <- df %>%
  select(sigma, epsilon, alfa, sigma_comp, assymetry,
         avg_shannon, avg_bbp, alive_predators, mean_tl, mean_n_of_species)

# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "complete.obs")

# Plot correlation matrix
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black",
         number.cex = 0.7, mar = c(0, 0, 1, 0))

print(cor_matrix)
```

## Extracting highly correlated pairs
```{r}
# Extracting pairs with high correlation (absolute value > 0.7)
high_corr <- which(abs(cor_matrix) > 0.7 & abs(cor_matrix) < 1, arr.ind = TRUE)
if (nrow(high_corr) > 0) {
  cat("Highly correlated variable pairs (|correlation| > 0.7):\n")
  for (i in 1:nrow(high_corr)) {
    var1 <- rownames(cor_matrix)[high_corr[i, 1]]
    var2 <- colnames(cor_matrix)[high_corr[i, 2]]
    corr_value <- cor_matrix[high_corr[i, 1], high_corr[i, 2]]
    if (high_corr[i, 1] < high_corr[i, 2]) {
      cat(sprintf("%s and %s: %.2f\n", var1, var2, corr_value))
    }
  }
}
```

```{r, warning=FALSE, message=FALSE}
ggpair = ggpairs(numeric_vars,
                 lower = list(continuous = wrap("points", size = 0.5)),
                 upper = list(continuous = wrap("cor")),
                 diag = list(continuous = wrap("barDiag")))
ggpair
```

# Scatter plots for key relationships
## Highly correlated
### Average Competition Strength Vs Average BBP
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = sigma_comp, y = avg_bbp)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average Competition Strength Vs Average BBP",
       x = "Average Competition Strength", y = "Average BBP") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Average BBP vs Mean Trophic Level
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = avg_bbp, y = mean_tl)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average BBP Vs Mean Trophic Level",
       x = "Average BBP", y = "Mean Trophic Level") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

## Parameters Vs Metrics
### Sigma Vs Average Shannon
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = sigma, y = avg_shannon)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Sigma Vs Average Shannon Index",
       x = "Sigma (Interaction Strength)", y = "Average Shannon Index") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Sigma Vs Average Shannon by Epsilon
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = sigma, y = avg_shannon)) +
  theme_classic() +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~epsilon) +
  labs(title = "Average Shannon Index Vs Sigma",
       x = "Sigma (Interaction Strength)", y = "Average Shannon Index") +
  geom_smooth(method = "lm", color = "red")
```

### Average bbp Vs Epsilon
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = epsilon, y = avg_bbp)) +
  geom_point(alpha = 0.3) +
  xlim(c(0, 5)) +
  labs(title = "Epsilon Vs Average BBP",
       x = "Epsilon (Predator-Prey Asymmetry)", y = "Average BBP") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Sigma Vs Mean Trophic Level
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = sigma, y = mean_tl)) +
  geom_point(alpha = 0.3) +
  xlim(c(0, 5)) +
  labs(title = "Sigma Vs Mean Trophic Level",
       x = "Sigma (Average Interaction Strength)", y = "Mean Trophic Level") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

## Metrics Vs Metrics
### Average Shannon Vs Alive Predators
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = avg_shannon, y = alive_predators)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average Shannon Index Vs Alive Predators",
       x = "Average Shannon Index", y = "Proportion of alive predators") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Average Shannon Vs Average BBP
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = avg_shannon, y = avg_bbp)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average Shannon Index Vs Average BBP",
       x = "Average Shannon Index", y = "Average BBP") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

**But if we look at Mean Number of Species (Average Species Richness) instead of the Shannon Index, then we see a clearer relationship.**

### Mean Species Richness Vs Average BBP
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = mean_n_of_species, y = avg_bbp)) +
  geom_point(alpha = 0.3, size = 0.8) +
  facet_wrap(~sigma_comp) + 
  labs(title = "Mean Species Richness Vs Average BBP",
       x = "Mean Species Richness", y = "Average BBP") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Average BBP Vs Alive Predators
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = avg_bbp, y = alive_predators)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average BBP Vs Alive Predators",
       x = "Average BBP", y = "Proportion of alive predators") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Average Species Richness Vs Mean Trophic Level
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = mean_n_of_species, y = mean_tl)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average Species Richness Vs Mean Trophic Level",
       x = "Average Species Richness", y = "Mean Trophic Level") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

**Here we see that high regional richness can only be maintained by reducing the amount o
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = mean_n_of_species, y = mean_tl, color = alive_predators)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Average Species Richness Vs Mean Trophic Level",
       x = "Average Species Richness", y = "Mean Trophic Level") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Alive Predators Vs Mean Trophic Level
```{r, warning=FALSE, message=FALSE}
ggplot(df, aes(x = mean_tl, y = alive_predators)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Alive Predators Vs Mean Trophic Level",
       x = "Mean Trophic Level", y = "Proportion of alive predators") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

### Biomass Pyramid Slope
```{r}
slope_df = read_csv("pyramid_slope_from_DirectSampling.csv")
```
**Merging slope to the original df**
```{r}
df_with_slope <- merge(df, slope_df, by = c("sigma", "epsilon", "alfa", "sigma_comp", "assymetry"))
```

```{r, warning=FALSE, message=FALSE}
ggplot(df_with_slope, aes(x = sigma, y = slope)) +
  facet_wrap(~epsilon) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Sigma Vs Biomass Slope",
       x = "Sigma (Interaction Strength)", y = "Biomass Pyramid Slope") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

```{r, warning=FALSE, message=FALSE}
ggplot(df_with_slope, aes(x = sigma_comp, y = slope)) +
  geom_point(alpha = 0.3, size = 0.8) +
  labs(title = "Sigma Vs Biomass Slope",
       x = "Sigma_comp (Competition Strength)", y = "Biomass Pyramid Slope") +
  geom_smooth(method = "lm", color = "red") +
  theme_classic()
```

# Linear Models
```{r}
slope_model_with_interaction = lm(slope ~ (sigma + epsilon + alfa + sigma_comp + assymetry)^2, data = df_with_slope)
summary(slope_model_with_interaction)
```
```{r}
# Backward elimination based on AIC
model_step <- step(slope_model_with_interaction, direction = "backward")

model_step
```

```{r}
# Define parameters
K <- 1
X <- seq(0, 2, by = 0.01)
Y <- X * (K - X)

# Plot the equation
plot(X, Y, type = "l", lwd = 2, col = "blue",
     main = expression(paste("Plot of ", X, "(", K, " - ", X, ")")),
     xlab = "X",
     ylab = expression(X(K - X)),
     ylim = c(-2, max(Y)))

# Add grid lines for better readability
grid()
```
```{r}
# Install and load the deSolve package if not already installed
if (!require(deSolve)) {
  install.packages("deSolve")
  library(deSolve)
}

# Define the ODE function
logistic_ode <- function(time, state, parameters) {
  with(as.list(c(state, parameters)), {
    dX <- X * (K - X)
    list(c(dX))
  })
}

# Parameters
parameters <- c(K = 1)

# Initial state
state <- c(X = 0.1)

# Time sequence
times <- seq(0, 10, by = 0.01)

# Solve the ODE
out <- ode(y = state, times = times, func = logistic_ode, parms = parameters)

# Convert output to data frame
out_df <- as.data.frame(out)

# Calculate dX/dt
out_df$dX_dt <- with(out_df, X * (parameters["K"] - X))

# Plot Population and dX/dt over Time
library(ggplot2)

# Melt the data for easier plotting
library(reshape2)
melted_df <- melt(out_df, id.vars = "time", measure.vars = c("X", "dX_dt"),
                  variable.name = "Variable", value.name = "Value")

# Create the plot
ggplot(melted_df, aes(x = time, y = Value, color = Variable)) +
  geom_line(size = 1) +
  labs(title = expression(paste("Population Dynamics: ", frac(dX, dt), " = X(K - X)")),
       x = "Time",
       y = "Value",
       color = "Variable") +
  theme_minimal() +
  ylim(c(0, 1.2)) +
  theme(plot.title = element_text(hjust = 0.5))

```



















